---
title: "Binding, Merging, and Reshaping"
format:
  html:
    toc: true
    toc-location: left
    theme: vapor
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

### Binding

Binding is generally a pretty easy operation, but it is best done when you have the same number of elements in the vectors that you are trying to bind together.  If you don't, you will get a warning message and the shorter vector will be recycled to match the length of the longer vector.

There are two types of binds: row binds and column binds.  Row binds are done with `rbind()` and column binds are done with `cbind()`.  The `r` and `c` stand for row and column, respectively.

```{r}
# Create some vectors
x <- 1:5
y <- 6:10
z <- 11:15

# Row bind
rbind(x, y, z)

# Column bind
cbind(x, y, z)
```

You'll see that the elements `x`, `y`, and `z` are the same, but how we bind them together changes the resultant object. When we row bind, we get a matrix with 3 rows and 5 columns.  When we column bind, we get a matrix with 5 rows and 3 columns.

What you'll usually find most handy is cbinding a vector to a data frame.

```{r}
# Create a data frame
test_df <- data.frame(
  x = 1:5, 
  y = 6:10, 
  z = 11:15
  )

# Create a vector
test_vec <- 1:5

# Column bind the vector to the data frame
cbind(test_df, test_vec)
```

### Merging

The great thing about modern languages is that they allow you to have multiple objects in your environment -- there is nothing stopping your from having 10 data frames in your environment.  However, there are times when you want to combine data frames together.  This is where merging comes in. Merging is a way to combine data frames together based on a common variable or variables. The base `merge()` function is used to merge data frames together and it is pretty easy to use. The main arguments are:

- `x` and `y`: the data frames you want to merge together
- `by`: the variable(s) that you want to merge on
- `all`: whether or not you want to keep all of the observations in the data frames you are merging together.  The default is `FALSE`, which means that only observations that are in both data frames will be kept.  If you set this to `TRUE`, then all observations will be kept, even if they are not in both data frames.

The `by` and `all` arguments can also be specified as `by.x` and `by.y` and `all.x` and `all.y`. The two `by` variants are useful if the variable you want to merge on has different names in the data frames you are merging together. The all variants are useful if you want to keep all observations in one data frame, but only some observations in the other data frame.

We can translate the all arguments into SQL words:

- all.x = TRUE & all.y = FALSE: left join
- all.x = FALSE & all.y = TRUE: right join
- all.x = TRUE & all.y = TRUE: full join
- all.x = FALSE & all.y = FALSE: inner join

### Reshaping

You will find yourself reshaping data a lot.  The most common reshaping operation is converting data from wide format (how you usually encounter data) to long format (sometimes called key-value pairs). There are a lot of functions that can do this work, but we will start with `reshape2`. The main function in `reshape2` is `melt()`, which takes your data from wide to long.  The main arguments are:

- `data`: the data frame you want to reshape
- `id.vars`: the variables you want to keep as columns
- `measure.vars`: the variables you want to convert to rows
- `variable.name`: the name of the column that will contain the variable names
- `value.name`: the name of the column that will contain the values

```{r}
library(reshape2)

# Create a data frame

test_df <- data.frame(
  id = 1:5,
  x = 1:5,
  y = 6:10,
  z = 11:15
  )

test_df
```

```{r}
# Melt the data frame

melt(
  data = test_df,
  id.vars = "id",
  measure.vars = c("x", "y", "z"),
  variable.name = "variable",
  value.name = "value"
  )
```

You can also go from long to wide with `dcast()`.  The main arguments are:

- `data`: the data frame you want to reshape
- `formula`: the formula that describes how to reshape the data
- `value.var`: the variable that contains the values

```{r}
# Create a data frame

test_df <- data.frame(
  id = c(1, 1, 2, 2, 3, 3),
  variable = c("x", "y", "x", "y", "x", "y"),
  value = c(1, 6, 2, 7, 3, 8)
  )

test_df  
```

```{r}
# Cast the data frame

dcast(
  data = test_df,
  formula = id ~ variable,
  value.var = "value"
  )
```

The formula is always a bit of a pain, so here is a quick guide:

- `id ~ variable`: the variable on the left side of the `~` is the variable that will be the rows and the variable on the right side of the `~` is the variable that will be the columns.

- `id ~ variable1 + variable2`: you can have multiple variables on the right side of the `~` and they will be combined to create the columns.

- `id ~ variable1 + variable2 + variable3`: you can have as many variables as you want on the right side of the `~`.

- `id ~ variable1 * variable2`: you can also combine variables on the right side of the `~` with `*`.  This will create a column for each combination of the variables.

- `id ~ variable1 + variable2 + variable3 + variable1:variable2`: you can also combine variables on the right side of the `~` with `:`.  This will create a column for each combination of the variables, but it will also create a column for each variable.

You can see why it can become a pain -- again, start with the easiest possible version and then work your way up. It is also never a bad idea for you to test with a smaller subset of the data, since these can get pretty computationally intensive.

## Practice